Evaluation of Pre-Trained Models for Many-to-English Translation

### 2024-08-19 paper/paper.tex
slight update
updated, now 16 pages

\documentclass[a4paper]{article}

@@

\begin{abstract}
    Machine translation (MT) has become an essential tool for overcoming language barriers in an increasingly globalised world. Today, there are pre-trained models (PTMs) specifically designed for automatic machine translation. This paper evaluates the performance of various pre-trained models (PTMs) for many-to-English translation. The models assessed include OPUS-MT \cite{tiedemann-2020-opus-mt}, mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}. Using a dataset curated from the Tatoeba repository, comprising 1,323 parallel sentence pairs across 14 source languages, translation quality is measured using BLEU, SacreBLEU, and METEOR scores. The results indicate that OPUS-MT, a one-to-one PTM, outperforms multilingual models in one-to-English translation tasks, with mBART-50 leading among the multilingual PTMs. These findings highlight the strengths and limitations of both specific language and multilingual PTMs in machine translation.
\end{abstract}

@@

According to Ethnologue \cite{ethnologue-2024}, 7,164 languages currently exist and in use today, with 40\% of them considered endangered. As of July 2024, 243 languages are supported by Google Translate (according to Wikipedia \cite{wikipedia-google-translate}). In modern times, the need for translation services has surged due to the growing exchange of information across different regions that speak various languages \cite{okpor-2014-machine-ta}.

Machine translation (MT) is the task of automatically translating from one language to another. This can be done through text or audio. It can be traced back to 1949 \cite{weaver-1999}, with the first public demonstration of an MT system on January 7, 1954, in collaboration with IBM, where 49 Russian sentences were translated into English using a limited vocabulary of 250 words and 6 grammar rules \cite{hutchins-2006-first-mt}. However, over the next several decades, growth were limited for machine translation, with 1956-1966 considered the decade of high expectation and disillusion, and 1967-1976 dubbed 'the quiet decade' \cite{hutchins-2001-mt-50-years}. Then in 1989, the dominance of the rule-based approach has been challenged by the rise of new methods and strategies, collectively referred to as ‘corpus-based’ methods (data-driven) \cite{hutchins-1994-research-methods-mt,hutchins-1998-development-mt}. Subsequently, statistics-based approaches for MT re-emerged, bolstered by the recent success of probabilistic techniques in speech recognition. Statistical machine translation \cite{lopez-2008-smt} dominated the domain between late 1990s through the early 2010s, before largely being surpassed by neural machine translation (NMT) \cite{cho-2014-properties,sutskever-2014-seq2seq}.

Since the introduction of Transformers in 2017 \cite{vaswani-2017-attention}, Natural Language Processing (NLP) and machine translation in particular reached a giant milestone. The following years saw the birth of Large Language Models (LLMs) such as BERT \cite{devlin-2019-bert}, GPT \cite{openai-2024-gpt4}, and T5 \cite{raffel-2023-t5}, which revolutionised both MT and the whole field of NLP. Then in early 2020s, several pre-trained models (PTM) that are specifically designed for machine translation emerged, OPUS-MT \cite{tiedemann-2020-opus-mt} produced numerous PTMs designed for one-to-one translations. Additionally, multilingual models: mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100} have also been published, growing popularity as multilinguality has been shown to allow for sharing information among languages \cite{garcia-2020-multilingual}. Most of these models are trained on diverse set of languages, allowing for many-to-many translation: able to translate between any of the supported pair of languages. This allows the models to generalise over shared lexical and linguistic among languages, and have been shown to increase performance compared to one-to-one translation models \cite{liu-2020-mbart}.

Despite these advancements, pre-trained models are often evaluated using different set of languages and benchmarks \cite{liu-2020-mbart,nllb200-2020,fan-2020-m2m100,wei-2023-polylm}, making it difficult to gauge their relative effectiveness across various languages. Nevertheless, there does not seem to be much work on comparing or benchmarking different pre-trained models in machine translation. Intento published 'The State of Machine Translation 2024' \cite{intento-2020} providing an in-depth evaluation of popular MT engines and LLMs. However, the biggest drawback in this report is that the selected LLMs are general LLMs such as GPT, LLaMa, Mistral, instead of MT-specific LLMs.

Through simple experimentations, this paper aims to evaluate the performance of existing pre-trained models (PTMs) on many-to-English translation across 14 source languages. Although fine-tuning multilingual PTMs has been proven to increase model performance \cite{cooper-stickland-2021-recipes}, no pre-training or fine-tuning is performed in this study for simplicity reasons. A dataset is curated from the Tatoeba repository \cite{tatoeba}, containing 1,323 parallel sentence pairs across source and target languages. The models include a one-to-one PTMs: OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt}, and multilingual PTMs: such as mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}. The performance of these models is evaluated using the BLEU \cite{papieni-2002-bleu}, SacreBLEU \cite{post-2018-sacrebleu}, and METEOR \cite{lavie-2007-meteor}.

The result indicates that OPUS-MT outperforms all other multilingual models in one-to-English translation, highlighting the superiority of specific languages PTMs compared to multilingual PTMs. mBART-50 shows the best performance among other multilingual PTMs: M2M-100 and NLLB-200.

@@

The encoder-decoder approach \cite{cho-2014-properties} remains as the foundation architecture for many sequence-to-sequence models in machine translation. Attention mechanism has been used for Neural Machine Translation (NMT) \cite{bahdanau-2016-nmt-jointly, luong-2015-effective-attention} before the release of the transformer \cite{vaswani-2017-attention}

Currently, Pre-Trained-Models (PTMs) have become a cornerstone in machine translation, leveraging large datasets and powerful architectures to achieve state-of-the-art performance \cite{han-2021-ptms}. Most PTMs support two kinds of translations: one-to-one translation and many-to-many translation (multilingual models).

One-to-One Translation refers to a translation approach where a model is specifically trained to translate between one source language and one target language. This setup is characterised by having a dedicated model for each unique language pair. OPUS-MT \cite{tiedemann-2020-opus-mt} and Facebook Fairseq \cite{ott-2019-fairseq} are examples of PTMs that support one-to-one translation. Multilingual models have also gained significant prominence, enabling many-to-many translation, where a single model can translate between multiple source and target languages \cite{aharoni-2019-massively-multilingual}.

OPUS-MT \cite{tiedemann-2020-opus-mt} is a project dedicated to creating and providing free resources for machine translation,providing over a thousand of PTMs of direct one-to-one translations. In contrast, mBART \cite{liu-2020-mbart}, M2M \cite{fan-2020-m2m100}, NLLB-200 \cite{nllb200-2020}, are massively multilingual models that support many-to-many translation language.

@@

Corpora are large and structured sets of texts used for linguistic research and analysis. Thus, parallel corpora are defined as sets of texts in a given source language along with their translations in another target language \cite{lefer-2020-parallel-corpora}. It can be bilingual or multilingual, and are crucial for training and evaluating machine translation. Languages without or with limited parallel corpora are referred as \textbf{low-resources language}.

Since the early 2000s, English has often been used as an intermediary language, meaning texts originally written in languages are first translated into English, and then from English into other languages, resulting in most target texts being translations of translations \cite{lefer-2020-parallel-corpora}. This is called indirect translation, and posed for many problems in translations as terms can be ambiguous with varying or similar meanings \cite{rosa-2017-indirect-translation-problems}.

Neural Machine Translation (NMT) systems require vast amounts of training data, and thus the availability of parallel corpora is crucial for building effective models \cite{koehn-2017-challenges}. The lack of extensive parallel corpora, especially for low-resource languages, leads to suboptimal performance in NMT techniques compared to their high-resource counterparts \cite{ranathunga-2023-nmt-low-res}. OPUS \cite{opus} is a comprehensive collection of open-source parallel corpora used extensively in the field of machine translation (MT). It includes corpora for 744 languages and contains over 1,210 different datasets, amassing a total of 45,945,946,108 sentence pairs. Tatoeba \cite{tatoeba} is another prominent resource in the field of MT and NLP, known for its extensive collection of translated sentences. As of July 2024, it contains 12,186,207 sentences over 423 supported languages, growing daily through volunteer contributions.

Several datasets are often used for evaluation in machine translation. The Tatoeba Challenge \cite{tiedemann-2020-tatoeba-challenge} covers 487 languages in 4,024 language pairs, including 657 test sets sourced from Tatoeba website, covering 138 languages. The TED Talks dataset \cite{ye-2018-word-embeddings-ted} contains transcripts from TED talks for more than 50 languages. IWSLT \cite{agarwal-iwstl-2023} also contains TED talks data, but paired with English translations. Finally, WMT \cite{barrault-2020-wmt} is an annual event that organises tasks for machine translation and provides a collection of datasets for benchmarking and evaluating translation systems.

While these datasets can be used to evaluate translation performance between any language pairs, they do not contain parallel corpora between different languages. Corpus between English and French for example, contains different text to the corpus between English and Spanish. Therefore, they do not support easy evaluation for many-to-English translations. Furthermore, papers often use different datasets and versions to evaluate their model performance, making direct comparison challenging.

mBart \cite{liu-2020-mbart} uses WMT19 for English-German translation and TED15 for Chinese-English translation. M2M-100 \cite{fan-2020-m2m100} uses 7 different datasets shared across languages. NLLB-200  \cite{nllb200-2020} uses FLORES dataset, designed for low-resources languages. OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt} models use datasets from Tatoeba challenges.

@@

Tatoeba is a vast, continuously expanding database consisting sentences and their translations, built through the contributions of thousands of volunteers, offering a tool that allows users to see examples of how words are used in sentences \cite{tatoeba}. They currently have 12,132,349 sentences and 423 supported languages, with around one to two thousand new sentences added daily, on average. The English sentence dataset contains 1,905,089 sentences, the largest one in their repository, with Russian in the second place with 1,066,633 sentences. Some languages supported on the website is shown in Figure \ref{fig:tatoeba_languages} and Figure \ref{fig:tatoeba_top_bottom_languages}, sorted from the biggest corpus.

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.9\linewidth]{images/tatoeba_languages.png}
    \caption{Tatoeba's languages repository with 10,000+ sentences and 100,000+ sentences \cite{tatoeba}}
    \label{fig:tatoeba_languages}
\end{figure}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.5\linewidth]{images/tatoeba_top_20_lang.png}
    \includegraphics[width=0.46\linewidth]{images/tatoeba_bottom_20_lang.png}
    \caption{Tatoeba top 20 and bottom 20 languages based on sentences count \cite{tatoeba}}
    \label{fig:tatoeba_top_bottom_languages}
\end{figure}

\begin{table}[htbp]
    \centering
    \large
    \begin{tabular}{|l|l|r|}
        \hline
        \textbf{No.} & \textbf{Language Pairs} & \textbf{Sentence Pairs Count} \\
        \hline
        1            & Chinese-English         & 68,814                        \\
        2            & Dutch-English           & 155,856                       \\
        3            & Finnish-English         & 102,202                       \\
        4            & French-English          & 405,088                       \\
        5            & German-English          & 501,145                       \\
        6            & Hebrew-English          & 172,082                       \\
        7            & Hungarian-English       & 171,698                       \\
        8            & Italian-English         & 624,160                       \\
        9            & Japanese-English        & 270,116                       \\
        10           & Polish-English          & 77,345                        \\
        11           & Russian-English         & 722,837                       \\
        12           & Spanish-English         & 265,253                       \\
        13           & Turkish-English         & 710,279                       \\
        14           & Ukrainian-English       & 214,244                       \\
        \hline
    \end{tabular}
    \caption{List of chosen languages for evaluation}
    \label{table:eval_languages}
\end{table}

Table \ref{table:eval_languages} show the 14 languages selected for this project. Languages are chosen based on its resources' availability in Tatoeba, as well as considering supported languages in most PTMs models. Thus, the languages chosen here can be considered as moderate to high resources languages.

To build the dataset, sentences in English are first downloaded, containing 1,898,494 sentences (it is unclear why it is less than the number stated in the Tatoeba website). Then for each language, sentence pairs between English and source languages are downloaded individually and compiled. The result is a single Dataframe containing 1,323 parallel sentences in all 14 languages, this will be treated as a test set to evaluate the models performance on each language. Note that in the case of multiple reference translations are available for the same sentence in English, only the first sentence is taken while the rest is discarded. Therefore, there is only one reference for each candidate sentence.

Sentences typically consist of everyday phrases such as 'I have to go to sleep', 'That is intriguing', and 'Where do you live?'. They may also include single-word exclamations like 'Speak!', 'So what?', or 'Look!'. Additionally, multiple sentences such as 'You may write in any language you want. On Tatoeba, all languages are considered equal', and 'Guns don't kill people. People kill people' can be found inside the corpus. A few of them also include human names, 'Compare your answer with Tom's', 'Muiriel is 20 now'. All of the sentences are straightforward and literal, without the use of linguistic features such as metaphors or sarcasm. Therefore, machine translation process should be straightforward on this level. Figure \ref{table:parallel_sentence} shows two examples of parallel sentences in the final dataset, across all 14 languages and the original English sentence.

\begin{CJK}{UTF8}{gbsn}
    \begin{table}[htbp]
        \centering
        \begin{tabular}{|l|l|l|}
            \hline
            \textbf{Language} & \textbf{Sentence 1}                   & \textbf{Sentence 2}                   \\
            \hline
            English           & I have to go to sleep.                & So what?                              \\
            Chinese           & 我该去睡觉了。                               & 那又怎樣?                                 \\
            Dutch             & Ik moet gaan slapen.                  & Dus?                                  \\
            Finnish           & Minun täytyy mennä nukkumaan.         & Mitä sitten?                          \\
            French            & Je dois aller dormir.                 & Et alors ?                            \\
            German            & Ich muss jetzt schlafen.              & Na und?                               \\
            Hebrew            & <hidden-due-to-latex-incompatibility> & <hidden-due-to-latex-incompatibility> \\
            Hungarian         & Aludni kell mennem.                   & És akkor mi van?                      \\
            Italian           & Devo andare a dormire.                & E allora?                             \\
            Japanese          & 私は眠らなければなりませ                          & だから何？                                 \\
            Polish            & Muszę iść spać.                       & No i co?                              \\
            Russian           & Мне пора идти                         & Так что?                              \\
            Spanish           & Tengo que irme a dormir.              & ¿Entonces qué?                        \\
            Turkish           & Yatmaya gitmek zorundayım.            & Öyleyse ne yapmalı?                   \\
            Ukrainian         & Маю піти спати.                       & Ну то що?                             \\
            \hline
        \end{tabular}
        \caption{A snippet of the dataset}
        \label{table:parallel_sentence}
    \end{table}
\end{CJK}

@@

Figure \ref{fig:word_count_box} shows a box plot of sentences word count. The median for Dutch, Finnish, French, German, Hebrew, Italian, Polish, Russian, Spanish, Turkish, Ukrainian seems relatively low, with a moderate spread. A few to moderate count of outlier exist in all languages, suggesting occasional longer sentences, with the highest variability seen in Japanese. Note that for Chinese and Japanese the count is based on each Chinese/Japanese letter, which explains their higher interquartile range (IQR) and larger numbers of outliers.

@@

\textbf{OPUS-MT} \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt} provides over 1,000 pre-trained models for translation between numerous language pairs. The architecture is based on MARIAN-NMT \cite{mariannmt}, based on a standard transformer setup: 6 self-attentive layers in both the encoder and decoder networks, each with 8 attention heads per layer \cite{tiedemann-2020-opus-mt}.

While they provide both monolingual and multilingual PTMs, only the OPUS-MT monolingual models are used in this study.

@@

\textbf{mBART} \cite{liu-2020-mbart} is a sequence-to-sequence denoising auto-encoder model specifically designed for multilingual tasks. The mBART-50 variant supports many-to-many translations for over 50 languages.

@@

The mBART model is built as a typical sequence-to-sequence transformer architecture \cite{vaswani-2017-attention}, using an embedding dimension of 1024 and 16 heads within 12 layers of encoder and 12 layers of decoder, totalling to around 680M parameters \cite{liu-2020-mbart}.

Figure \ref{fig:mbart} shows  mBART's key features: multilingual denoising pre-training and fine-tuning. The dataset is first corrupted by a noising function, which is applied to 35\% of words of each span text \cite{liu-2020-mbart}. During pre-training, the model will be made to denoise the texts, improving the model's ability to generate coherent and accurate translations. After the pre-training step, the model is then able to be fine-tuned for downstream tasks. It is shown that the model achieved consistent performance gains through pre-training in low-to-medium resource sentence level MT \cite{liu-2020-mbart}

@@

\textbf{M2M-100} \cite{fan-2020-m2m100} is designed to perform direct translation between any pair of 100 languages without relying on English as an intermediate language. Similarly, the architecture is of a transformer sequence-to-sequence with a typical encoder that outputs embeddings of the same length, while the decoder is an autoregressive component that sequentially produce the target sentence one token at a time. The parameters include 12 encoder layers and 12 decoder layers, 8192 FFN size and 1024 embedding dimension, totalling to a 1.2B parameter count \cite{fan-2020-m2m100}.

>>>>

\documentclass[a4paper, 11pt]{article}

@@

\begin{abstract}
    Machine translation (MT) has become an essential tool for overcoming language barriers in an increasingly globalised world. Today, there are pre-trained models (PTMs) specifically designed for automatic machine translation. This paper evaluates the performance of various pre-trained models (PTMs) for many-to-English translation. The models assessed include OPUS-MT \cite{tiedemann-2020-opus-mt}, mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}. Using a dataset curated from the Tatoeba repository, comprising 1,323 parallel sentence pairs across 14 source languages, translation quality is measured using BLEU, SacreBLEU, and METEOR scores. The results indicate that OPUS-MT, a one-to-one PTM, outperforms multilingual models in one-to-English translation tasks, with mBART-50 leading among the multilingual PTMs. Chinese, Italian, Polish, and Turkish translations to English show the best performance on average across the four models used in this study, however, Ukrainian-to-English on the OPUS-MT model achieves the highest SacreBLEU and METEOR scores.
\end{abstract}

@@

According to Ethnologue \cite{ethnologue-2024}, 7,164 languages currently exist and are in use today, with 40\% of them considered endangered. With so many languages around us, the language barrier has become a known issue, especially in the current globalised world where it is essential for individuals who speak different languages to be able to communicate effectively \cite{nakamura-2009-overcoming}. Thus, the need for translation services to facilitate communication across different languages has also surged \cite{okpor-2014-machine-ta}. As of August 2024, only 243 languages are supported by Google Translate (according to Wikipedia \cite{wikipedia-google-translate}).

Machine translation (MT) refers to the task of automatically translating from one language to another, which typically can be done through text or audio. The research and practice can be traced back to 1949 \cite{weaver-1999}, with the first public demonstration of an MT system on January 7, 1954, in collaboration with IBM, where 49 Russian sentences were translated into English using a limited vocabulary of 250 words and 6 grammar rules \cite{hutchins-2006-first-mt}. However, over the next several decades, growth was limited for machine translation, with 1956-1966 considered the decade of high expectation and disillusion, and 1967-1976 dubbed 'the quiet decade' \cite{hutchins-2001-mt-50-years}. Then in 1989, the dominance of the rule-based approach was challenged by the rise of new methods and strategies, collectively referred to as ‘corpus-based’ methods (data-driven) \cite{hutchins-1994-research-methods-mt,hutchins-1998-development-mt}. Subsequently, statistics-based approaches for MT re-emerged, bolstered by the recent success of probabilistic techniques in speech recognition. Statistical machine translation \cite{lopez-2008-smt} dominated the domain between the late 1990s through the early 2010s, before largely being surpassed by neural machine translation (NMT) \cite{cho-2014-properties,sutskever-2014-seq2seq}.

Since the introduction of Transformers in 2017 \cite{vaswani-2017-attention}, Natural Language Processing (NLP) and machine translation in particular reached a giant milestone. The following years saw the birth of Large Language Models (LLMs) such as BERT \cite{devlin-2019-bert}, GPT \cite{openai-2024-gpt4}, and T5 \cite{raffel-2023-t5}, which revolutionised both MT and the whole field of NLP. Then in the early 2020s, several pre-trained models (PTM) that are specifically designed for machine translation emerged, OPUS-MT \cite{tiedemann-2020-opus-mt} produced numerous PTMs designed for one-to-one translations. Additionally, multilingual models: mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100} have also been published, growing popularity as multilingualism has been shown to allow for sharing information among languages \cite{garcia-2020-multilingual}. Most of these models are trained on a diverse set of languages, allowing for many-to-many translation, which is able to translate between any of the supported pairs of languages. This allows the models to generalise over shared lexical and linguistic among languages, and have been shown to increase performance compared to one-to-one translation models \cite{liu-2020-mbart}.

Despite these advancements, pre-trained models are often evaluated using a different set of languages and benchmarks \cite{liu-2020-mbart,nllb200-2020,fan-2020-m2m100,wei-2023-polylm}, making it difficult to gauge their relative effectiveness across various languages. Nevertheless, there does not seem to be much work on comparing or benchmarking different pre-trained models in machine translation. Intento recently published 'The State of Machine Translation 2024' \cite{intento-2020} providing an in-depth evaluation of 52 MT engines and LLMs across 11 language pairs. Intento reported a lack of growth in the number of supported languages and concluded that translation quality significantly varies depending on languages \cite{intento-2020}. However, the biggest drawback in this report is that the selected LLMs are general LLMs such as GPT-4 \cite{openai-2024-gpt4}, LLaMa2 \cite{touvron-2023-llama2}, and Mistral \cite{jiang-2023-mistral-7b}, instead of MT-specific models.

Furthermore, BLEU \cite{papieni-2002-bleu} has been shown to be inconsistent depending on tokenisation and parameters that are used \cite{post-2018-sacrebleu}. While most researchers do report on these specifications explicitly, different BLEU parametrisations between different papers still make a direct comparison challenging to do.

This study will lean towards evaluating MT-specific models rather than general LLMs. Through simple experimentations, this paper aims to evaluate the performance of existing and publicly available MT pre-trained models (PTMs) on many-to-English translation across 14 source languages. Although fine-tuning multilingual PTMs has been proven to increase translation performance \cite{cooper-stickland-2021-recipes}, no pre-training or fine-tuning is performed in this study due to simplicity reasons and resource limitations. A dataset is curated from the Tatoeba repository \cite{tatoeba}, containing 1,323 parallel sentence pairs across source and target languages. The models include a one-to-one PTMs: OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt}, and multilingual PTMs: such as mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}. The performance of these models is evaluated using the BLEU \cite{papieni-2002-bleu}, SacreBLEU \cite{post-2018-sacrebleu}, and METEOR \cite{lavie-2007-meteor}.

The result indicates that OPUS-MT outperforms all other multilingual models in one-to-English translation on both SacreBLEU and METEOR scores, highlighting the superiority of specific languages PTMs compared to unfine-tuned multilingual PTMs. Among the multilingual models, mBART-50 shows the best performance, while NLLB-200 performs the worst, by a significant margin. Ukrainian-to-English translation by OPUS-MT receives the highest scores, however, Chinese, Italian, Polish, and Turkish are better on average across the four models.

@@

Currently, pre-trained models (PTMs) have become a cornerstone in machine translation, leveraging large datasets and powerful architectures to achieve state-of-the-art performance \cite{han-2021-ptms}. Attention mechanism has been used for Neural Machine Translation (NMT) \cite{bahdanau-2016-nmt-jointly, luong-2015-effective-attention} even before the introduction of the transformer \cite{vaswani-2017-attention}. The transformer, with its encoder-decoder approach \cite{cho-2014-properties}, remains the foundational architecture for many sequence-to-sequence models in machine translation. Most PTMs support two kinds of translations: one-to-one translation and many-to-many translation (multilingual models).

One-to-one translation refers to a translation approach where a model is specifically trained to translate between one source language and one target language. This setup is characterised by having a dedicated model for each unique language pair. OPUS-MT \cite{tiedemann-2020-opus-mt} is a project dedicated to creating and providing free resources for machine translation, providing over a thousand PTMs of direct one-to-one translations. Additionally, Facebook Fairseq \cite{ott-2019-fairseq} similarly offers one-to-one PTMs, although they are only limited to translations between English, French, German, and Russian.

Many-to-many translation allows for a translation between multiple pairs of source and target languages. This is done by training a single multilingual model, typically trained on multilingual datasets. Multilingual models have recently gained significant prominence, shown to be simpler and efficient \cite{aharoni-2019-massively-multilingual}. mBART \cite{liu-2020-mbart}, M2M \cite{fan-2020-m2m100}, NLLB-200 \cite{nllb200-2020}, are examples of massively multilingual models that support many-to-many translation language.

@@

Corpora are large and structured sets of texts used for linguistic research and analysis. Thus, parallel corpora are defined as sets of texts in a given source language along with their translations in another target language \cite{lefer-2020-parallel-corpora}. It can be bilingual or multilingual and is crucial for training and evaluating machine translation. Languages without or with limited parallel corpora are referred to as low-resource languages.

Since the early 2000s, English has often been used as an intermediary language, meaning texts originally written in languages are first translated into English, and then from English into other languages, resulting in most target texts being translations of translations \cite{lefer-2020-parallel-corpora}. This is called indirect translation and poses many problems in translations as terms can be ambiguous with varying or similar meanings \cite{rosa-2017-indirect-translation-problems}.

Neural Machine Translation (NMT) systems require vast amounts of training data, and thus the availability of parallel corpora is crucial for building effective models \cite{koehn-2017-challenges}. The lack of extensive parallel corpora, especially for low-resource languages, leads to suboptimal performance in NMT techniques compared to their high-resource counterparts \cite{ranathunga-2023-nmt-low-res}. OPUS \cite{opus} is a comprehensive collection of open-source parallel corpora used extensively in the field of machine translation (MT). It includes corpora for 744 languages and contains over 1,210 different datasets, amassing a total of 45,945,946,108 sentence pairs. Tatoeba \cite{tatoeba} is another prominent resource in the field of MT and NLP, known for its extensive collection of translated sentences. As of July 2024, it contains 12,186,207 sentences over 423 supported languages, growing daily through volunteer contributions. There are currently
86,915 members contributed through the Tatoeba website.

Several datasets are often used for evaluation in machine translation. The Tatoeba Challenge \cite{tiedemann-2020-tatoeba-challenge} covers 487 languages in 4,024 language pairs, including 657 test sets sourced from the Tatoeba website, covering 138 languages. The TED Talks dataset \cite{ye-2018-word-embeddings-ted} contains transcripts from TED talks for more than 50 languages. IWSLT \cite{agarwal-iwstl-2023} also contains TED talks data, but paired with English translations. Finally, WMT (Conference on Machine Translation) \cite{barrault-2020-wmt}, previously named Workshop on Statistical Machine Translation, is an annual event that organises tasks for machine translation and provides a collection of datasets for benchmarking and evaluating translation systems.

While these datasets can be used to evaluate translation performance between any language pairs, they do not contain parallel corpora between multiple different languages. A corpus between English and French, for example, contains different sets of text than a corpus between English and Spanish. Therefore, they do not support easy evaluation for many-to-English translations. Furthermore, papers often use different datasets and versions to evaluate their model performance, making direct comparisons challenging. mBart \cite{liu-2020-mbart} uses WMT19 for English-German translation and TED15 for Chinese-English translation. M2M-100 \cite{fan-2020-m2m100} uses seven different datasets shared across languages. NLLB-200  \cite{nllb200-2020} uses the FLORES dataset, designed for low-resources languages. OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt} models are evaluated using the datasets from Tatoeba challenges.

@@

The Tatoeba website \cite{tatoeba} includes a vast, continuously expanding database consisting of sentences and their translations, built through the contributions of thousands of volunteers. They currently have 12,132,349 sentences of over 423 supported languages, with around one to two thousand new sentences added daily, on average. The English sentence dataset contains 1,905,089 sentences, the largest one in their repository, with Russian in second place with 1,066,633 sentences. Figure \ref{fig:tatoeba_languages} shows the repository, sorted from the biggest corpus.

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.65\linewidth]{images/tatoeba_languages.png}
    \includegraphics[width=0.34\linewidth]{images/tatoeba_top_20_lang.png}
    \caption{Tatoeba's repository by languages, sorted by sentences count \cite{tatoeba}}
    \label{fig:tatoeba_languages}
\end{figure}

\begin{table}[htbp]
    \centering
    \large
    \begin{tabular}{llr}
        \hline
        \textbf{No.} & \textbf{Language Pairs} & \textbf{Sentence Pairs Count} \\
        \hline
        1            & Chinese-English         & 68,814                        \\
        2            & Dutch-English           & 155,856                       \\
        3            & Finnish-English         & 102,202                       \\
        4            & French-English          & 405,088                       \\
        5            & German-English          & 501,145                       \\
        6            & Hebrew-English          & 172,082                       \\
        7            & Hungarian-English       & 171,698                       \\
        8            & Italian-English         & 624,160                       \\
        9            & Japanese-English        & 270,116                       \\
        10           & Polish-English          & 77,345                        \\
        11           & Russian-English         & 722,837                       \\
        12           & Spanish-English         & 265,253                       \\
        13           & Turkish-English         & 710,279                       \\
        14           & Ukrainian-English       & 214,244                       \\
        \hline
                     & Total                   & 4,461,119                     \\
        \hline
    \end{tabular}
    \caption{List of chosen languages for evaluation}
    \label{table:eval_languages}
\end{table}

Table \ref{table:eval_languages} shows the 14 language pairs selected for this study. Languages are chosen based on the resources' availability in Tatoeba and supported languages in most PTMs. Thus, the languages chosen here can be considered moderate to high-resource languages.

To build the dataset, 1,898,494 English sentences were initially downloaded, though the number is less than what is listed on the Tatoeba website for reasons that remain unclear. For each of the 14 source languages, sentence pairs in English were individually downloaded and compiled. The English corpus was then merged with each compiled source language corpus to create parallel sentences across 14 languages. In each corpus, multiple translations for the same English phrase were dropped, retaining only the first translation to ensure that each English sentence had only a single translation reference. The final dataset contains 1,323 parallel sentences in English and all 14 source languages. This corpus will be used as a test set to evaluate the models' performance in each language.

Sentences typically consist of everyday phrases such as 'I have to go to sleep', 'That is intriguing', and 'Where do you live?'. They may also include single-word exclamations like 'Speak!', 'So what?', or 'Look!'. Additionally, multiple sentences such as 'You may write in any language you want. On Tatoeba, all languages are considered equal', and 'Guns don't kill people. People kill people' can be found inside the corpus. A few of them also include human names, 'Compare your answer with Tom's', and 'Muiriel is 20 now'. All the sentences seem to be straightforward and literal, without any use of linguistic features such as metaphors or sarcasm. Figure \ref{table:parallel_sentence} shows a couple of examples of parallel sentences in the final dataset, across all 14 languages and the original English sentence.

\begin{CJK}{UTF8}{gbsn}
    \begin{table}[htbp]
        \centering
        \begin{tabular}{lrr}
            \hline
            \textbf{Language} & \textbf{Sentence 1}                   & \textbf{Sentence 2}                   \\
            \hline
            English           & I have to go to sleep.                & So what?                              \\
            Chinese           & 我该去睡觉了。                               & 那又怎樣?                                 \\
            Dutch             & Ik moet gaan slapen.                  & Dus?                                  \\
            Finnish           & Minun täytyy mennä nukkumaan.         & Mitä sitten?                          \\
            French            & Je dois aller dormir.                 & Et alors ?                            \\
            German            & Ich muss jetzt schlafen.              & Na und?                               \\
            Hebrew            & <hidden-due-to-latex-incompatibility> & <hidden-due-to-latex-incompatibility> \\
            Hungarian         & Aludni kell mennem.                   & És akkor mi van?                      \\
            Italian           & Devo andare a dormire.                & E allora?                             \\
            Japanese          & 私は眠らなければなりませ                          & だから何？                                 \\
            Polish            & Muszę iść spać.                       & No i co?                              \\
            Russian           & Мне пора идти                         & Так что?                              \\
            Spanish           & Tengo que irme a dormir.              & ¿Entonces qué?                        \\
            Turkish           & Yatmaya gitmek zorundayım.            & Öyleyse ne yapmalı?                   \\
            Ukrainian         & Маю піти спати.                       & Ну то що?                             \\
            \hline
        \end{tabular}
        \caption{A snippet of the dataset}
        \label{table:parallel_sentence}
    \end{table}
\end{CJK}

@@

Figure \ref{fig:word_count_box} shows a box plot of each language sentence's word count. Note that for Chinese and Japanese the word count is counted for each Chinese/Japanese letter instead, which explains their higher range and larger numbers of outliers.

Most of the languages' interquartile range (IQR) is quite low (around two) with a centre median, except for Chinese, Italian, and Japanese. Finnish and Hebrew have the lowest median, while Polish, Russian, Spanish, Turkish, and Ukrainian have a median closely similar to each other. English, Dutch, French, German, and Italian have a slightly higher median, with Italian also having a wider IQR. Lastly, Japanese and Chinese have the highest median, the widest range, and significantly more outliers.

@@

OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt} provides over 1,000 pre-trained models for translation between numerous language pairs. The architecture is based on MARIAN-NMT \cite{mariannmt}, based on a standard transformer setup: 6 self-attentive layers in both the encoder and decoder networks, each with 8 attention heads per layer \cite{tiedemann-2020-opus-mt}.

While the framework provides both monolingual and multilingual PTMs, only the OPUS-MT monolingual models are used in this study. This is partly due to the lack of documentation on the multilingual models.

@@

mBART \cite{liu-2020-mbart} is a sequence-to-sequence denoising auto-encoder model specifically designed for multilingual tasks. The mBART-50 variant supports many-to-many translations for over 50 languages.

@@

The mBART model is built as a typical sequence-to-sequence transformer architecture \cite{vaswani-2017-attention}, using an embedding dimension of 1024 and 16 heads within 12 layers of encoder and 12 layers of decoder, totalling around 680M parameters \cite{liu-2020-mbart}.

Figure \ref{fig:mbart} shows mBART's key features: multilingual denoising pre-training and fine-tuning. The dataset is first corrupted by a noising function, which is applied to 35\% of words of each span text \cite{liu-2020-mbart}. During pre-training, the model will be made to denoise the texts, improving the model's ability to generate coherent and accurate translations. After the pre-training step, the model is then able to be fine-tuned for downstream tasks. It is shown that the model achieved consistent performance gains through pre-training in low-to-medium resource sentence-level translation \cite{liu-2020-mbart}

@@

M2M-100 \cite{fan-2020-m2m100} is designed to perform direct translation between any pair of 100 languages without relying on English as an intermediate language. Similarly, the architecture is of a transformer sequence-to-sequence with a typical encoder that outputs embeddings of the same length, while the decoder is an autoregressive component that sequentially produces the target sentence one token at a time. The parameters include 12 encoder layers and 12 decoder layers, 8192 hidden layer sizes, and 1024 embedding dimensions, totalling a 1.2B parameter count \cite{fan-2020-m2m100}.
